---
title: "Week 03 Lab: Tidying ASEAN Meat-Production Data"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(readxl)
library(waldo)
```


## 3.1 Summary of Team Predictions (Sheet X.14)

As the Team Reporter, I have compiled the predictions from Jun Lum, Benjamin, Bryan, and Zen regarding the tidying of sheet X.14. 

### Part A & B: Columns, Classes, and Row Counts

The table below summarizes the team's expectations for the final tidy dataset structure.

| Member | Expected Columns & Classes | Expected Rows | Reasoning / Formula |
| :--- | :--- | :--- | :--- |
| **Jun Lum** | Country (Char), Animal Type (Char), Year (Num), Production (Num) | 160 | $10 \text{ countries} \times 4 \text{ years} \times 4 \text{ animal types}$ |
| **Benjamin** | Country (Char), Animal Type (Char), Year (Num), Production (Num) | 160 | $10 \text{ countries} \times 4 \text{ years} \times 4 \text{ animal types}$ |
| **Bryan** | Country (Char), Animal Type (Char), Year (Num), Production (Num) | 160 | $10 \text{ countries} \times 4 \text{ years} \times 4 \text{ animal types}$ |
| **Zen** | Country (Char), Animal Type (Char), Year (Num), Production (Num) | 160 | $10 \text{ countries} \times 4 \text{ years} \times 4 \text{ animal types}$ |

---

### Part C: Why is Sheet X.14 not tidy

The team identified several distinct violations of tidy data principles in the current layout:

*   **Structural Misplacement of Variables:** 
    *   **Years:** Jun Lum and Zen identified years acting as column headers, whereas ben and Bryan noted that years are serving as section labels/stacked within rows. 
    *   **Headers:** Jun Lum noted individual country names are being used as column headers.
*   **Mixing Data Levels (Raw vs. Aggregated):** Jun Lum, Benjamin, and Zen all noted that summary columns/rows (e.g., the **ASEAN total column**) are mixed directly into the raw data.
*   **Data Quality & Redundancy:** 
    *   Benjamin pointed out that "Poultry" rows contain no data and should be removed entirely.
    *   Bryan noted that missing values are represented inconsistently using symbols like `-`.

---

### Reporter summary (team)

- **Agreed tidy variables:** `country` (character), `animal` (character), `year` (numeric), `production` (numeric).
- **Agreed row count:** 160 rows ($10 \times 4 \times 4$).
- **Agreed non-tidy issues:** header-encoded years, layout-encoded variables, and mixing summary totals (ASEAN/Total) with raw observations.
- **Discrepancies to resolve during cleaning:** Ben expects poultry to have no data and be removable; others treat poultry as one of the four animal categories required by the task. Bryan notes missing values shown as `-` in the source.

## 3.2 Import the ASEAN Meat Production Data

Open `ASEAN-Statistical-Yearbook-2023.xlsx` and identify the cell range containing the table in sheet `X.14`.

```{r}
#| label: import-meat
meat_raw <- read_xlsx(
  path = "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "X.14",
  range = "A4:L36"
)

meat_raw
```

## 3.3 Tidy the Meat Production Data

```{r}
#| label: tidy-meat
meat_tidy <- meat_raw %>%
  rename(label = 1) %>%
  select(where(~ !all(is.na(.x)))) %>%
  select(-any_of("ASEAN")) %>%
  mutate(
    label = str_squish(as.character(label)),
    year = as.numeric(if_else(
      str_detect(label, "^\\d{4}$"),
      label,
      NA_character_
    ))
  ) %>%
  fill(year, .direction = "down") %>%
  mutate(
    label_lower = str_to_lower(label),
    animal = case_when(
      str_detect(label_lower, "beef") &
        str_detect(label_lower, "buffalo") ~ "beef and buffalo",
      str_detect(label_lower, "^pig") ~ "pig",
      str_detect(label_lower, "mutton") |
        str_detect(label_lower, "goat") ~ "mutton and goat",
      str_detect(label_lower, "^poultry") &
        !str_detect(label_lower, "production") ~ "poultry",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-label_lower) %>%
  filter(!is.na(animal), year %in% 2018:2021) %>%
  pivot_longer(
    cols = -c(label, year, animal),
    names_to = "country",
    values_to = "production",
    values_transform = list(production = as.character)
  ) %>%
  mutate(
    country = str_squish(country),
    production = readr::parse_number(production, na = c("-", ""))
  ) %>%
  arrange(desc(year), animal, country) %>%
  select(country, animal, year, production)

meat_tidy %>% slice_head(n = 6)
meat_tidy %>% slice_tail(n = 6)

# Verifier checks
nrow(meat_tidy)
meat_tidy %>% distinct(animal) %>% arrange(animal)
```

## 3.4 Correct Cambodia’s Mutton-and-Goat Entries

Justification (brief): we target exactly the four Cambodia + mutton-and-goat rows where the XLSX encodes missing “-” as numeric 0 for years 2018–2021, and replace only those values with `NA`.

```{r}
#| label: cambodia-fix
meat_tidy <- meat_tidy %>%
  mutate(
    production = if_else(
      country == "Cambodia" &
        animal == "mutton and goat" &
        year %in% 2018:2021 &
        production == 0,
      NA_real_,
      production
    )
  )

meat_tidy %>%
  filter(country == "Cambodia", animal == "mutton and goat") %>%
  arrange(desc(year))
```

## 3.5 Checkpoint: Compare `meat_tidy` to Reference Data

```{r}
#| label: compare-meat
meat_ref <- read_csv("meat_tidy_reference.csv", show_col_types = FALSE)
compare(meat_tidy, meat_ref, tolerance = 1e-6)
```

## 3.6 Tidy the Population Data

Import sheet `I.1` (population in thousands), excluding the ASEAN aggregate row.

```{r}
#| label: import-pop
pop_raw <- read_xlsx(
  path = "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "I.1",
  skip = 4
)

pop_raw
```

```{r}
#| label: tidy-pop
pop_tidy <- pop_raw %>%
  rename(country = 1) %>%
  select(where(~ !all(is.na(.x)))) %>%
  mutate(country = str_squish(as.character(country))) %>%
  filter(!is.na(country), country != "", country != "ASEAN") %>%
  filter(
    !str_detect(country, "^Source"),
    !str_detect(country, "ASEAN Secretariat")
  ) %>%
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "pop"
  ) %>%
  mutate(
    year = as.numeric(year),
    pop = as.numeric(pop)
  ) %>%
  filter(year %in% 2013:2022) %>%
  arrange(desc(year), country) %>%
  select(country, year, pop)

pop_tidy %>% slice_head(n = 6)

# Verifier note: read_xlsx() trims whitespace by default (trim_ws = TRUE),
# so a trailing space in "Malaysia " is typically removed during import.
```

## 3.7 Checkpoint: Compare `pop_tidy` to Reference Data

```{r}
#| label: compare-pop
pop_ref <- read_csv("population_tidy_reference.csv", show_col_types = FALSE)
compare(pop_tidy, pop_ref, tolerance = 1e-6)
```

## 3.8 Join Meat Production to Population

```{r}
#| label: join
merged <- meat_tidy %>%
  left_join(pop_tidy, by = c("country", "year")) %>%
  select(country, animal, year, production, pop)

# Verifier: ensure row ordering matches meat_tidy
identical(
  merged %>% select(country, animal, year),
  meat_tidy %>% select(country, animal, year)
)
```

## 3.9 Derive Per-Capita Production

Production is in thousand metric tons; population is in thousand persons.
Thus $(\text{production}/\text{pop})$ is metric tons per person; multiply by 1000 to get kg per person.

```{r}
#| label: per-capita
merged <- merged %>%
  mutate(meat_per_cap = production / pop * 1000)

merged %>% slice_head(n = 6)
```

## 3.10 Checkpoint: Compare `merged` to Reference Data

```{r}
#| label: compare-merged
merged_ref <- read_csv("merged_reference.csv", show_col_types = FALSE)
compare(merged, merged_ref, tolerance = 1e-6)
```

## 3.11 Visualize the Result

```{r}
#| label: plot
ggplot(
  merged,
  aes(x = year, y = meat_per_cap, color = animal)
) +
  geom_line() +
  geom_point() +
  labs(
    x = NULL,
    y = "Production (kg per person)",
    color = "Animal",
    title = "Per-Capita Meat Production in ASEAN, 2018–2021",
    caption = "Source: ASEAN Statistical Yearbook 2023"
  ) +
  xlim(2017.9, 2021.1) +
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0.01", "0.1", "1", "10", "100")
  ) +
  facet_wrap(~country) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 0),
    legend.justification.inside = c(1, 0.05)
  )
```

## 3.12 Comment on the Plot

**Noteworthy Feature:** One striking feature in the plot is the sharp decline in **pig meat production** (the teal line) in the **Philippines** between 2019 and 2021. While the production of poultry and beef in the country remains relatively flat and stable on the log scale, per-capita pig production drops significantly during this two-year window.

**Interpretation & Source:** This steep, isolated decline is the result of the **African Swine Fever (ASF)** outbreak. ASF was first confirmed in the Philippines in the second half of 2019. The highly contagious viral disease decimated local swine populations, leading to massive culling efforts by the government and a drastic reduction in domestic pork output over the following years. *(Source: USDA Foreign Agricultural Service reports on Philippine Livestock; FAO situation updates on ASF in Asia).*

## 4.1 Reflect on AI Use

- Tool used: (fill in)
- One useful AI suggestion: (fill in)
- One AI response that was wrong/incompatible/etc. and how the team corrected it: (fill in)

## 4.2 Roles and Contributions

- Member A (Coder): Chue Zen Whye
- Member B (Navigator + Verifier): Chan Jun Lum
- Member C (Constructive Critic + Reporter): Lim Wei Zhe Bryan
- Member D (AI Liaison): Tan Jie Ming. Benjamin
