---
title: "Week 03 Lab: Tidying ASEAN Meat-Production Data"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(readxl)
library(waldo)
```

## 3.1 Think Before Prompting (Team Predictions)

### a) Columns and classes

| Column | Class | Example |
|---|---|---|
| country | character | "Brunei Darussalam" |
| animal | character | "beef and buffalo" |
| year | numeric | 2021 |
| production | numeric | 4.26741 |

### b) Expected number of rows

We expect:

$10 \text{ countries} \times 4 \text{ years} \times 4 \text{ animal types} = 160$ rows.

Rationale: each observation corresponds to one country–animal–year combination with one production value.

### c) Why sheet X.14 is not tidy

- Years are used as column headers rather than values in a single `year` column.
- The layout uses country names / animal groupings as structural labels rather than explicit variables stored in columns.
- The sheet mixes raw observations with redundant aggregate information (e.g., ASEAN aggregate column and total rows).

## 3.2 Import the ASEAN Meat Production Data

Open `ASEAN-Statistical-Yearbook-2023.xlsx` and identify the cell range containing the table in sheet `X.14`.

```{r}
#| label: import-meat
# TODO: Replace `range` with the exact rectangle containing the table.
meat_raw <- read_xlsx(
  path = "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "X.14",
  range = "A1:Z99" 
)

meat_raw
```

## 3.3 Tidy the Meat Production Data

```{r}
#| label: tidy-meat
# TODO: Adjust the renaming/selection logic to match the actual column names in `meat_raw`.
# The goal is: country, animal, year, production.

meat_tidy <- meat_raw %>%
  # Common pattern for this sheet:
  # 1) An 'animal' label column that needs fill-down
  # 2) A 'country' column
  # 3) Year columns 2018-2021 that need pivot_longer
  rename_with(~ str_trim(.x)) %>%
  rename_with(~ str_replace_all(.x, "\\s+", " ")) %>%
  rename_with(tolower) %>%
  # If needed, rename the first two columns to animal/country:
  rename(
    animal = 1,
    country = 2
  ) %>%
  fill(animal) %>%
  mutate(
    animal = str_to_lower(animal)
  ) %>%
  pivot_longer(
    cols = -c(country, animal),
    names_to = "year",
    values_to = "production"
  ) %>%
  mutate(
    year = as.numeric(year),
    production = as.numeric(production)
  ) %>%
  # Keep only the required animals (exact strings) and years:
  filter(year %in% 2018:2021) %>%
  mutate(
    animal = recode(
      animal,
      "beef and buffaloes" = "beef and buffalo",
      .default = animal
    )
  ) %>%
  filter(animal %in% c("beef and buffalo", "mutton and goat", "pig", "poultry")) %>%
  # Drop redundant aggregate/total entries (may need adjustment based on actual sheet text):
  filter(country != "ASEAN") %>%
  filter(country != "Total meat production") %>%
  arrange(desc(year), animal, country) %>%
  select(country, animal, year, production)

meat_tidy %>% slice_head(n = 6)
meat_tidy %>% slice_tail(n = 6)

# Verifier checks
nrow(meat_tidy)
meat_tidy %>% distinct(animal) %>% arrange(animal)
```

## 3.4 Correct Cambodia’s Mutton-and-Goat Entries

Justification (brief): we target exactly the four Cambodia + mutton-and-goat rows where the XLSX encodes missing “-” as numeric 0 for years 2018–2021, and replace only those values with `NA`.

```{r}
#| label: cambodia-fix
meat_tidy <- meat_tidy %>%
  mutate(
    production = if_else(
      country == "Cambodia" &
        animal == "mutton and goat" &
        year %in% 2018:2021 &
        production == 0,
      NA_real_,
      production
    )
  )

meat_tidy %>%
  filter(country == "Cambodia", animal == "mutton and goat") %>%
  arrange(desc(year))
```

## 3.5 Checkpoint: Compare `meat_tidy` to Reference Data

```{r}
#| label: compare-meat
meat_ref <- read_csv("meat_tidy_reference.csv", show_col_types = FALSE)
compare(meat_tidy, meat_ref, tolerance = 1e-6)
```

## 3.6 Tidy the Population Data

Open the XLSX and identify the cell range containing the table in sheet `I.1` (exclude the ASEAN aggregate row).

```{r}
#| label: import-pop
# TODO: Replace `range` with the exact rectangle containing the table.
pop_raw <- read_xlsx(
  path = "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "I.1",
  range = "A1:Z99"
)

pop_raw
```

```{r}
#| label: tidy-pop
pop_tidy <- pop_raw %>%
  rename_with(~ str_trim(.x)) %>%
  rename_with(~ str_replace_all(.x, "\\s+", " ")) %>%
  rename_with(tolower) %>%
  rename(country = 1) %>%
  filter(country != "ASEAN") %>%
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "pop"
  ) %>%
  mutate(
    year = as.numeric(year),
    pop = as.numeric(pop)
  ) %>%
  arrange(desc(year), country) %>%
  select(country, year, pop)

pop_tidy %>% slice_head(n = 6)

# Verifier note: read_xlsx() trims whitespace by default (trim_ws = TRUE),
# so a trailing space in "Malaysia " is typically removed during import.
```

## 3.7 Checkpoint: Compare `pop_tidy` to Reference Data

```{r}
#| label: compare-pop
pop_ref <- read_csv("population_tidy_reference.csv", show_col_types = FALSE)
compare(pop_tidy, pop_ref, tolerance = 1e-6)
```

## 3.8 Join Meat Production to Population

```{r}
#| label: join
merged <- meat_tidy %>%
  left_join(pop_tidy, by = c("country", "year")) %>%
  select(country, animal, year, production, pop)

# Verifier: ensure row ordering matches meat_tidy
identical(
  merged %>% select(country, animal, year),
  meat_tidy %>% select(country, animal, year)
)
```

## 3.9 Derive Per-Capita Production

Production is in thousand metric tons; population is in thousand persons.
Thus $(\text{production}/\text{pop})$ is metric tons per person; multiply by 1000 to get kg per person.

```{r}
#| label: per-capita
merged <- merged %>%
  mutate(meat_per_cap = production / pop * 1000)

merged %>% slice_head(n = 6)
```

## 3.10 Checkpoint: Compare `merged` to Reference Data

```{r}
#| label: compare-merged
merged_ref <- read_csv("merged_reference.csv", show_col_types = FALSE)
compare(merged, merged_ref, tolerance = 1e-6)
```

## 3.11 Visualize the Result

```{r}
#| label: plot


ggplot(
  merged,
  aes(x = year, y = meat_per_cap, color = animal)
) +
  geom_line() +
  geom_point() +
  labs(
    x = NULL,
    y = "Production (kg per person)",
    color = "Animal",
    title = "Per-Capita Meat Production in ASEAN, 2018–2021",
    caption = "Source: ASEAN Statistical Yearbook 2023"
  ) +
  xlim(2017.9, 2021.1) +
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0.01", "0.1", "1", "10", "100")
  ) +
  facet_wrap(~country) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 0),
    legend.justification.inside = c(1, 0.05)
  )
```

## 3.12 Comment on the Plot

(Write 1+ noteworthy feature and cite a credible source that helps interpret it.)

## 4.1 Reflect on AI Use

- Tool used: (fill in)
- One useful AI suggestion: (fill in)
- One AI response that was wrong/incompatible/etc. and how the team corrected it: (fill in)

## 4.2 Roles and Contributions

- Member A (Coder): …
- Member B (Navigator + Verifier): …
- Member C (Constructive Critic + Reporter): …
- Member D (AI Liaison): …
